The promise of community-driven preprints in ecology and evolution

### 2024-05-04 ms/ms.qmd
Some text changes, some bigger comments to change yet. Including re-organising percentages

##---------------------##
# Types pf preprints
##---------------------##
# Summary of the data
  sum_types <- preprints  %>% tabyl(type_of_preprint)  %>% rename("type" = type_of_preprint, "n" = n)  %>% mutate(perc = n/sum(n)*100, labels = paste0(round(perc, 1), "%"))
  sum(sum_types$n) #620
  p4 <- ggplot(sum_types, aes(x = "", y = perc, fill = type)) +
    geom_col(color = "black") +
    geom_label_repel(aes(label = labels),
              position = position_stack(vjust = 0.6),
            show.legend = FALSE) +
    scale_fill_viridis_d() +
    coord_polar(theta = "y") + 
    theme_void() + labs(fill = "Article Type") + theme(plot.tag = element_text(size = 16, face = "bold"), plot.tag.location = "margin")

@@

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| fig-cap: Summary of the publication status of preprints on *EcoEvoRxiv*.  A) Time between submission and publication of preprints on *EcoEvoRxiv*. 'Preprints' that were published within 2 months (60 days) of being submitted are considered postprints. B) Number of preprints that were published in open access journals. C) Open access status of preprints published on *EcoEvoRxiv*.

@@

- First, authors are encouraged to share their data and code at the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. Reassuringly, sharing data and code is rarely associated with 'scooping' of research findings [@soeharjono2021reported]. If authors are worried about data being used unintentionally, clear information surrounding its reuse can include in a license (see https://choosealicense.com). Data can also be achived with an embargo on its reuse [@roche2014troubleshooting]. 
- Second, authors need to be aware of the benefits of posting preprints and postprints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible, it is still seems rare that constructive discussions form around preprints in an open forum (e.g. bioRxiv [@anderson2020biorxiv]). Unfortunately, the *EcoEvoRxiv* website does not provide opportunities for discussion given the limitations of the web server at this point in time. As such, we could not accurately assess how much discourse around a given preprint occurs. Clearly, as a community, we need to provide better platforms that document discussions around preprint findings. Such discussions help the authors improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing services, such as *Peer Community In* (PCI) or *PubPeer* (see also [@avissar2024recommendations] for more options), to provide feedback on preprints.
- Third, it remains clear that the time between posting a preprint and publication is still quite long. One possible explanation is that preprints are not being sent to suitable journals. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI provides authors with quicker reviews of a paper and a more realistic recommendation of where such work is likely to be published. Many journals now also accept PCI reviews and recommendations when publishing a paper. As such, we would encourage more authors to consider sending their preprints for PCI review to possibly speed up the publication process. 
- Finally, while most preprints get seamlessly connected with and merged with their published version some remain disconnected. Incorrect cross-linking can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are well understood and easily rectified. They often result from a mismatch between preprint and published metadata (e.g., titles and author details). In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). We found that mismatched metadata almost always contributed to preprints and published articles not being matched automatically in Google Scholar. We recommend that authors ensure that, when preprints are published in a journal, they update their preprint with the publication DOI. Especially if their title has changed, this is quite easy for authors to do on *EcoEvoRxiv*, and it will ensure that the preprint is linked to the published article and that the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the research findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv*, much work needs to be done to improve our community’s understanding and use of preprints (and post-prints). We view this perspective piece as one small step in achieving that goal. Authors must also be aware of the benefits of using community-driven preprint servers and the unique initiatives being pursued. Community-driven preprint servers are capable of setting their own agenda and are driven by the needs and desires of the community. Supporting these initiatives should be a priority for all researchers. Volunteers at *EcoEvoRxiv* should remain open to new and innovative ways to improve publication and open science practices. We believe that the future of preprints is bright, and community-driven initiatives, such as *EcoEvoRxiv*, are set to play a crucial role in the future of scientific publishing.

>>>>

##---------------------##
# Types pf preprints
##---------------------##
# Summary of the data
  sum_types <- preprints  %>% tabyl(type_of_preprint)  %>% rename("type" = type_of_preprint, "n" = n)  %>% mutate(perc = n/sum(n)*100, labels = paste0(round(perc, 1), "%"))
  #sum(sum_types$n) #620
  p4 <- ggplot(sum_types, aes(x = "", y = perc, fill = type)) +
    geom_col(color = "black") +
    geom_label_repel(aes(label = labels),
              position = position_stack(vjust = 0.6),
            show.legend = FALSE) +
    scale_fill_viridis_d() +
    coord_polar(theta = "y") + 
    theme_void() + labs(fill = "Article Type") + theme(plot.tag = element_text(size = 16, face = "bold"), plot.tag.location = "margin")

@@

# Following the journey of a preprint on *EcoEvoRxiv*: from submission to publication
```{r, fig-pubsummary}
#| label: fig-pubsummary
#| echo: false
#| warning: false
#| message: false
#| fig-cap: Summary of the publication status of preprints on *EcoEvoRxiv*.  A) Time between submission and publication of preprints on *EcoEvoRxiv*. 'Preprints' that were published within 2 months (60 days) of being submitted are considered postprints. B) Number of preprints that were published in open access journals. C) Open access status of preprints published on *EcoEvoRxiv*.

@@

- First, authors are encouraged to share their data and code at the preprint stage. Sharing data and code early can help improve the quality of research, establish precedence, and improve the transparency and computational reproducibility of scientific findings [@sarabipour2019value]. Reassuringly, sharing data and code is rarely associated with 'scooping' of research findings [@soeharjono2021reported]. If authors are worried about data being used unintentionally, clear information surrounding its reuse can include in a license (see https://choosealicense.com). Data can also be achived with an embargo on its reuse [@roche2014troubleshooting]. 
- Second, authors need to be aware of the benefits of posting preprints and postprints for constructive feedback [@bourne2017ten]. While it is clear preprints help establish precedence and allow findings to be openly accessible, it is still seems rare that constructive discussions form around preprints in an open forum (e.g. bioRxiv [@anderson2020biorxiv]). Unfortunately, the *EcoEvoRxiv* website does not provide opportunities for discussion given the limitations of the web server at this point in time. As such, we could not accurately assess how much discourse around a given preprint occurs. Clearly, as a community, we need to provide better platforms that document discussions around preprint findings. Such discussions help the authors improve their work and communicate their findings more effectively (when done constructively, of course). One way to facilitate such discussions may be to use open peer reviewing services, such as *Peer Community In* (PCI) or *PubPeer* (see also [@avissar2024recommendations] for more options), to provide feedback on preprints.
- Third, it remains clear that the time between posting a preprint and publication is still quite long. One possible explanation is that preprints are not being sent to suitable journals. Interestingly, out of the `r nrow(data2)` preprints, only `r pci[2,2]` (`r pci[2,3]*100`%) used Peer Community In (PCI) as a preprint peer-review service. Using PCI circumvents editorial decisions without review. As such, it ensures that authors receive faster feedback on a paper and a more realistic recommendation of where such work is likely to be published. Many journals now accept PCI reviews and recommendations when publishing a paper, which may help facilitate faster publication; though more journals could incorperate PCI in thier policies. 
- Finally, while most preprints get seamlessly connected and merged with their published version some remain disconnected. Incorrect cross-linking by indexing platforms (e.g., Google Scholar) can create confusion and lead to frustration among authors. Thankfully, the reasons for unmatched preprints and publications are well understood and easily rectified. They often result from a mismatch between preprint and published metadata (e.g., titles and author details). In fact, it was reasonably common for titles of preprints to change between the preprint and the published article (`r titles[2,2]` or `r titles[2,3]*100`%). We found that mismatched metadata almost always contributed to preprints and published articles not being matched automatically in Google Scholar. We recommend that authors ensure that, when preprints are published in a journal, they update their preprint with the publication DOI, especially if their title has changed. This is quite easy for authors to do on *EcoEvoRxiv*, and it will ensure that the preprint is linked to the published article and that the citations are merged. It will also create a link between the final preprint and the published paper so that, no matter the open access status of the publication, the research findings will remain openly accessible, saving costs for authors and improving the visibility and use of research.

Despite the early successes of the new initiatives taken by *EcoEvoRxiv*, much work needs to be done to improve our community’s understanding and use of pre- and post-prints. We view this perspective piece as a small step towards achieving that goal. Authors must also be aware of the benefits of using community-driven preprint servers and the unique initiatives being pursued. Community-driven preprint servers are capable of setting their own agenda and are driven by the needs and desires of the community. Supporting these initiatives should be a priority for all researchers. Volunteers at *EcoEvoRxiv* are encouraged to remain open to new and innovative ways to improve publication and open science practices. We believe that the future of preprints is bright, and community-driven initiatives, such as *EcoEvoRxiv*, are set to play a crucial role in the future of scientific publishing.
